---
title: "02-qPCR-data-plates1-2"
output: html_notebook
---

24 January 2023


After running the rest of the current batch of samples, let's take a look at the qPCR results and temporal trends.

Because many of the samples amplify at a higher number of cycles than my standards, the appropriate thing to do would be to re-run the samples with a new set of standards that overlap more with the cycle numbers for the samples (i.e., drop the highest concentration standards in favor of additional standards on the lower end).

For the meantime, the preliminary data show a clear temporal trend at the ADN site, but lower overall detections (and conc. of detections) at the MCB site. It would be good to know more about those...




```{r load-libraries}
library(tidyverse)
library(readxl)
```


```{r load-data}
test_data <- read_xlsx("../data/2023-01-21_WinterRunChinook_testplate_export.xlsx", sheet = "Results", skip = 34)

# plate 1
plate1 <- read_xlsx("../data/2023-01-23_winterRunChinook_plate1.xlsx", sheet = "Results", skip = 32)

# plate 2
plate2 <- read_xlsx("../data/2023-01-23_WinterRunChinook_plate2.xlsx", sheet = "Results", skip = 32)
```



Merge that data into a single df.

```{r}
data.qpcr <- test_data %>%
  bind_rows(., plate1, plate2) %>% 
  # remove the IPC test data since we went with just the Chinook probe
  filter(!str_detect(`Sample Name`, "IPC")) %>%
  select(`Sample Name`, Task, CT, `Ct Mean`, `Ct SD`, Quantity, `Quantity Mean`, `Quantity SD`)


```


```{r}
data.qpcr %>%
  filter(Task != "STANDARD",
         CT != "Undetermined") %>%
  ggplot(aes(x = `Ct Mean`, y = log10(`Quantity Mean`))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Cycle (Ct) vs. log DNA copy number/ul",
       x = "Cycle (Ct)",
       y = "log Quantity (DNA copy number/ul)")

```
The relationship drifts at higher cycle numbers - that could be a plate effect?

Samples amplified on multiple plates?

```{r}
sample.data <- data.qpcr %>%
   filter(Task != "STANDARD", `Sample Name` != "NTC") %>%
  group_by(`Sample Name`) %>%
  add_tally() %>%
  arrange(`Sample Name`) 

# look at amplification per sample 
sample.data %>%
  filter(n > 3) %>%
   ggplot(aes(x = `Sample Name`, y = Quantity)) +
  geom_point() +
  theme_minimal() +
  labs(title = "McCloud eDNA qPCR test - DNA copy number/ul per sample",
       x = "Sample",
       y = "Mean quantity (DNA copy number/ul)") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
  )
  
```
Cycle number for that subset of samples?

```{r}
sample.data %>%
  filter(n > 3, 
         CT != "Undetermined") %>%
  ggplot(aes(x = CT, y = log10(Quantity))) +
  geom_point() +
  geom_text(aes(label = `Sample Name`), check_overlap = T, size = 3) +
  theme_minimal() +
  labs(title = "Cycle (Ct) vs. log DNA copy number/ul",
       x = "Cycle (Ct)",
       y = "log Quantity (DNA copy number/ul)")
  
```



Quickly look at trends in sample means over the temporal period sampled.

```{r}
sample.data %>%
 ggplot(aes(x = `Sample Name`, y = `Quantity Mean`)) +
  geom_point() +
  theme_minimal() +
  labs(title = "McCloud eDNA qPCR test - DNA copy number/ul per sample",
       x = "Sample",
       y = "Mean quantity (DNA copy number/ul)") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
  )


```

Make two factors for the sites: ADN and MCB

```{r}
# this plot does not use the mean DNA copy number quantity - so there should be multiple data points per sample replicate
sample.data %>%
  mutate(site = ifelse(str_detect(`Sample Name`, "ADN"), "ADN", "MCB")) %>%
  ggplot(aes(x = `Sample Name`, y = Quantity, color = site)) +
  geom_point(alpha = 0.7) +
  theme_bw() +
  facet_grid(cols = vars(site), scales = "free_x") +
  labs(title = "McCloud eDNA qPCR: DNA copy number/ul, replicates",
       x = "Sample",
       y = "Quantity (DNA copy number/ul)") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
  )

ggsave("pdf_outputs/plates1_2_output_by_replicate.pdf", height = 5, width = 8)
```

Do the quick conversion to get to DNA copies per L (rather than ul).


```{r}
sample.data %>%
  mutate(mean_quantity_L = `Quantity Mean`*10^6) %>%
  mutate(site = ifelse(str_detect(`Sample Name`, "ADN"), "ADN", "MCB")) %>%
  ggplot(aes(x = `Sample Name`, y = mean_quantity_L, color = site)) +
  geom_point() +
  theme_bw() +
  facet_grid(cols = vars(site), scales = "free_x") +
  labs(title = "McCloud eDNA qPCR test - mean DNA copy number/L",
       x = "Sample",
       y = "Mean quantity (DNA copy number/L)") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
  )

```

## Add metadata

```{r}
meta <- read_xlsx("../data/WRCSMcCloud22_eDNAMetadata_CJS_dsb.xlsx", sheet = "meta_simple")

meta 
```

combine qPCR and metadata
```{r}
merged.df <- sample.data %>%
  left_join(., meta, by = c("Sample Name" = "Sample ID")) %>%
  mutate(quantity_L = Quantity*10^6) %>%
  mutate(site = ifelse(str_detect(`Sample Name`, "ADN"), "ADN", "MCB"))


# plot with that info
merged.df %>%
  ggplot(aes(x = `Date Collected`, y = quantity_L, color = site)) +
  geom_point(alpha = 0.7) +
  theme_bw() +
  facet_grid(cols = vars(site), scales = "free_x") +
  labs(title = "McCloud eDNA qPCR: DNA copy number/L, by collection date",
       x = "Collection date, 2022",
       y = "Quantity (DNA copy number/L)") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
  ) 
  
ggsave("pdf_outputs/plates1_2_output_by_date.pdf", height = 5, width = 8)
```

